---
title: "Microbial Communities Metadata Statistical Analysis"
author: "Samantha Gallegos Perez"
date: "2026-03-11"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(car)
library(nlme)
library(emmeans)
```

# Load the Data

Loaded the data into a variable called soil_initial and used head() to check variables.
```{r load data}
soil_initial <- read.csv("trimmed_metadata.csv")#load the data
head(soil_initial)
```

## Question: Did fire management practice affect the relative abundance of bacteria, and does this depend on season?

##Hypothesis: I expect a higher bacterial relative abundance in the warmer months, with summer having the highest, followed by spring, and winter to have the lowest counts. I also predict bacterial amplicon counts will decrease immediately following fire regardless of season.

# Data Collection and Data Wrangling 

This metadata contains information of soil samples collected from remnant, restored, and  agricultural fields from ten sites at Nachusa Grasslands. This data set includes repeated measures completed on May 2 (Spring), August 8(Summer), and October 13 (Autumn) in 2017 (Mason et al., 2023). Additionally, fire management practices (y/n) were also included in this experiment. This study encouraged my interest to understand the role season and fire management practices play in the increase of relative abundance of bacterial counts using repeated studies. 

To test the hypothesis, the only variables, Site, Fire, bacteria, archaea, fungi and season. The continuous dependent variable would be bacteria, and independent variables would be Site, Fire, and Season. This data is a repeated measures design, making Fire and Season to be treated as fixed factors and Site as a random factor, making this a mixed model analysis. 

Before proceeding with analysis, the data was wrangled by converting our Fire, Season and Site variables from data type "char" to "factor". The original data set was filtered to only include Site, Fire, bacteria, archaea, fungi, and season. Additionally, the data was mutated to calculate relative abundance of bacterial counts by dividing bacterial values by the sum of bacteria, archaea, and fungi. 
```{r data wrangling}

#Only selected necessary columns that include soil, Site, Date, Type, fungi, bacteria and archaea 
filtered_soil <- select(soil_initial, Site, Date, Fire, fungi, bacteria, archaea, season)


#check for balance
table(filtered_soil$Site) #every site has three rows
table(filtered_soil$Fire)#yes to added fire has 12 values, and no to added fire has 18 values
#unbalanced
table(filtered_soil$season)#three equal amounts of 10

#these are bacterial counts, divided bacteria to the total of bacterial+archaea+fungal to get the relative abundance
final_soil <- mutate(filtered_soil, bact_relative_abund = bacteria/(bacteria+archaea+fungi))

#our random factor is Site, and fixed factors will be Fire and season
#changed "char" data type for season, Site, and fire to "factor"
final_soil$Site <- as.factor(final_soil$Site)
final_soil$season <- as.factor(final_soil$season)
final_soil$Fire <- as.factor(final_soil$Fire)

```

# Initial Inspection
A histogram was created to idenitify how the values of our dependent variable bact_relative abund look. The histogram looks slighly left-skewed. To add, an initial ggplot was created too. 
```{r initial plot}
#created a histogram 
png(filename = "Histogram_Plot.png")
hist(final_soil$bact_relative_abund)
dev.off()

#create an initial plot to demonstrate the how high the bacterial relative abundance levels are 
initial_plot <- ggplot(final_soil, aes(x=season,y=bact_relative_abund,fill=Fire)) +
  geom_boxplot() +
  scale_y_continuous(limits=c(0,1))

ggsave(filename = "initial_ggplot.png",initial_plot)#save and export the .png file
```



# Mixed Model Testing
The mixed model was created using the lme() function with bact_relative_abund as the  dependent variable, Season (fixed), Fire (fixed) and Site (random) as the independent variables. The residuals of the model, soil_model, were plotted to test assumptions for heteroskedascity and normality. The Anova() function was used to test the model's fixed factors with a type III Chi squared test.
```{r mixed model}

#created the linear mixed effect model including the IVs: Season and Fire, and DV:bact_relative_abund using the random variable as Site
soil_model <- lme(bact_relative_abund~season*Fire, random = ~1| Site, data = final_soil, method = "ML")

#exported the initial plot of residuals to Model_Residuals_Plot.png
png(filename = "Model_Residuals_Plot.png")
plot(soil_model)
dev.off()

#saved the qqnorm/qqline plot to Normal_Q-Q_plot.png
png(filename = "Normal_Q-Q_plot.png")
qqnorm(resid(soil_model))#initial inspections a linear line/checked the residuals for normality
qqplot_1 <- qqline(resid(soil_model))
dev.off()

#because the experiment was unbalanced, type III was used with the interaction and would need to change the contrasts before and after running Anova
options(contrasts = c("contr.sum","contr.poly"))
soil_model <- lme(bact_relative_abund~season*Fire, random = ~1| Site, data = final_soil, method = "ML") #remember to refit the model, and to reset contrast after
Anova(soil_model, type=3) #type 3 because there's an interaction
options(contrasts = c("contr.treatment","contr.poly"))
```

The summary() and emmeans() functions were used for post-hoc comparisons of the means. 
```{r summary}
#re-ran  model in order to get the summary 
soil_model <- lme(bact_relative_abund~season*Fire, random = ~1| Site, data = final_soil, method = "ML")
summary(soil_model)

#check the emmeans to analyze means from the summary 
emmeans(soil_model, specs=c('season','Fire'))
```


# Final Visualization
```{r final visualization}
#changed the order of the values of the season variable for visualization purposes
final_soil$season <- factor(final_soil$season, levels = c("Spring", "Summer", "Autumn"))

#create the final visualization
final_plot <- ggplot(final_soil, aes(x=season,y=bact_relative_abund,fill=Fire)) +
  geom_boxplot() +#use a box-plot for two categorical IVs, and 1 DV
  scale_fill_manual(values = c("y" = "orange", "n" = "lightblue" )) + #customize color:orange for fire(y), light blue for no fire(n)
  labs(title ="Effect of Fire and Season on Bacterial Relative Abundance", x ="Season", y="Bacterial Relative Abundance") +#created the titles 
  geom_point(position = position_dodge(width = .75)) +
  theme_classic() + #created a white background
  theme(plot.title = element_text(hjust = 0.5, size= 14), axis.title.x= element_text(size=12), axis.title.y=element_text(size=12), axis.text.x = element_text(size=11), axis.text.y = element_text(size=11),legend.title = element_text(size=14),legend.text = element_text(size=10))
#adjusted sizes of all text and titles within theme

#save the visualization
ggsave(filename = "Fire_Management_Season_on_Plot.png", final_plot)
```
# Conclusion
Based on the Analysis of Deviance Table, there was no significant interaction between fire management and season and how it affects the relative abundance of bacteria (χ = 1.04; df=2; χ ²= 0.592). Independently, Season (χ = 1.35; df=2; χ ²= 0.509) and Fire (χ = 1.40; df=1; χ ²= 0.237) did not show any significant affect on relative abundance either. Bacterial relative abundance was high and was not dependent on Season Fire, or their interaction. With added fire management practices, there was a 0.05% increase in abundance for spring, 0.09% for Summer and 0.17% for Autumn. 



